A White Noise Approach to Evolutionary Ecology

### 2020-04-13 .A White Noise Approach To Evolutionary Ecology/draft.tex
local lipschitz condition in main text

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued stochastic processes. This approach allows us to compute the full implications of demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a quantitative genetic model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests selection gradients can be either positively or negatively correlated with competition coefficients depending on the ratio of interspecific trait variation to intraspecific trait variation. Hence, this approach contributes to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of stochastic models that underlie rigorous investigations of the relationship between feedbacks of biological processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document: 
    extra_dependencies:
      - float
      - flafter
    fig_caption: yes
    fig_height: 3
    fig_width: 4
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
#knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
```

# Introduction

Our goal in this manuscript is to develop a rigorous, but accessible approach to synthesize the stochastic dynamics of abundance, mean trait and heritable variation in biological populations for the study of theoretical evolutionary ecology. A primary aim of theoretical evolutionary ecology is the development of mathematical approaches to describe the evolution of populations and their interactions with both the biotic and abiotic environments in which they are embedded. Given this consideration, a natural scope for such an approach centers on quantifying the abundance dynamics of populations and the evolution of traits mediating their interactions as functions of relevant abiotic factors. Although taking into account abundance, phenotype and environment provides the basis for a partial understanding of the complex nature of biological communities, a deeper understanding must account for the effects of dispersal and the phylogeographic history of interacting lineages [@Kraft2007; @Hickerson2010; @Manceau2016; @markmcpeek2017] along with the genetic basis of ecologically relevant traits [@jeffreyconner2004; @FUSSMANN2007] and feedbacks between populations and the biogeochemical cycles they ultimately depend on [@michelloreau2010; @9781107648258]. It is therefore ideal that the development of any such mathematical approach anticipates extensions to account for these important factors shaping ecological communities, especially as empirical and conceptual work in these directions continues to grow [@AbdalaRoberts2014; @Klzsch2015; @Crutsinger2015; @Fitzpatrick2015; @Fitzpatrick2017; @Marx2017; @Rudman2017; @Skovmand2018; @VanNuland2019; @Harmon2019]. Furthermore, the approach would benefit from a stochastic component to capture the chance nature of biological reality [@Lande2003; @DeMeester2018; @Mubayi2019] and serve as a basis for the construction of statistical methods that measure evolutionary and ecological processes occuring in the wild. Such methods will tether theory to reality and allow for rigorous tests of hypotheses on the structure and behavior of ecological communities. In this paper we introduce a framework that establishes a formal connection between the continuous-time dynamics of abundance and quantitative traits in stochastically evolving populations. We then demonstrate the utility of our framework through the derivation and analysis of a model of diffuse coevolution and discuss how it can be extended to account for the details mentioned above.

Current theoretical approaches to synthesize evolution and ecology have capitalized on the fact that biological fitness plays a key role in determining both sets of dynamics. While correlation of fitness and genotype is the basis of evolution by natural selection, the mean fitness across all individuals in a population determines the growth, stasis or decline of abundance. In section \ref{deterministic} we review the mathematical formalization of this connection, which has been established in the contexts of population genetics [@9781932846126; @joanroughgarden1979], evolutionary game theory [@Hofbauer1998; @9780674023383; @Lion2018], quantitative genetics [@Lande1982; @Doebeli1996; @Lion2018] and a unifying framework for these three distinct approaches to evolutionary theory [@Champagnat2006] which includes our approach as a special case. We note here there are two primary differences between our aims and those of Champagnat, Ferrière, and Méléard [-@Champagnat2006]. First, instead of developing a unifying approach to evolutionary theory, we focus on developing a stochastic synthesis of population dynamics and quantitative genetics including specific expressions for the dynamics of abundance, mean trait and additive genetic variance. Second, although Champagnat, Ferrière, and Méléard [-@Champagnat2006] provide rigorous proofs of results relevant to our approach, they do not communicate the nessecary tools we make use of for deriving the dynamics of abundance, mean trait and additive genetic variance. We aim to communicate these tools to a wide audience by providing an intuitive, yet accurate treatment of white noise and associated heuristics useful for deriving new stochastic models of evolutionary ecology.

Reviewing the above accomplishments reveals a beautiful synthesis of evolution and population ecology. However, it also reveals a gap in theoretical approaches to incorporate the intrinsically random nature of populations. Specifically, in theoretical quantitative genetics the derivation of a population's response to random genetic drift is derived in discrete time under the assumption of constant effective population size using arguments based on properties of random samples [@Lande1976]. Though this approach conveniently mimics the formalism provided by the Wright-Fisher model of population genetics, real population sizes fluctuate over time. Furthermore, since these fluctuations are themselves stochastic, it seems natural to derive expressions for the evolutionary response to demographic stochasticity and consider how the results relate to characterizations of random genetic drift. This can be done in continuous time for population genetic models without too much technical overhead, assuming a finite number of alleles [@Lande2009; @Parsons2010; @Gomulkiewicz2017]. However, for populations with a continuum of types, such as a quantitative trait, finding a formal approach to derive the evolutionary response to demographic stochasticity has remained a vexing mathematical challenge. Previous work integrating stochastic processes with quantitative genetic models have profited from the simplifying assumption that phenotypic and genetic variances remain constant over time, yeilding analytical approximations for the joint dynamics of abundance and the mean of a quantitative trait [@Engen2013]. These results not only ignore the evolution genetic variance, but they focus on environmental stochasticity and assume sufficiently large populations so that demographic stochasticity is negligible. Here we account for both demographic stochasticity and the evolution of additive genetic variance by combining the calculus of white noise with results on rescaled limits of branching Brownian motion processes (BBM) and stochastic partial differential equations (SPDE), which are stochastic analogs of deterministic partial differential equations (PDE). Our goal has two components: 1) Establish a novel synthetic approach to theoretical evolutionary ecology that provides a formal connection between demographic stochasticity and random genetic drift in the context of quantitative traits. 2) Communicate some useful properties of space-time white noise, BBM and SPDE to as wide of audience as possible. With this goal in mind we will not provide a rigorous treatment of any of these deep subjects. Instead, we introduce a set of heuristics that only require the basic concepts of Reimann integration, partial differentiation and some exposure to Brownian motion and stochastic ordinary differential equations (SDE). For a concise introduction to SDE and Brownian motion, we recommend the primer by Evans [-@9781470410544]. Rigorous treatments of SPDE and rescaled limits of BBM can be found in Walsh [-@Walsh] and Dawson [-@dawson1993measure] respectively.

Through accomplishing these goals this work yields five main results. Firstly, we develop a concise, but accurate approach for understanding SPDE and provide heuristics to empower mathematical evolutionary ecologists to derive novel models that relax key assumptions commonly made in combining models of evolutionary and abundance dynamics. Second, we formally derive general SDE that express the joint dynamics of abundance, mean trait and additive genetic variance for populations exhibiting a wide range of potential phenotypic distributions. Third, we simplify these SDE by approximating the abundance density of phenotypes with a Gaussian curve and integrate the results with a model of imperfect inheritance to find formula that generalize classical quantitative genetic expressions. This includes an expression for additive genetic variance analogous to the expression for the response in mean phenotype found by Lande [-@Lande1976]. Fourth, following classical niche theory we provide a rigorous mechanistic derivation of fitness for populations experiencing resource competition and abiotic stabilizing selection using diffusion limits of measure-valued branching processes. This calculation provides a demonstration for performing such diffusion limits that can be applied in other contexts. Fifth, we explore implications of the resulting model of diffuse coevolution by investigating the relationship between selection gradients and competition coefficients.

We now summarize the basic structure of the paper. To provide motivation for the stochastic equations developed later and background for our model of coevolution, we begin with \S\ref{deterministic} by briefly summarizing derivations of deterministic dynamics of biological populations. Starting with a PDE (which we refer to as the Deterministic Asexual Gaussian allelic model with Abundance dynamics, abbreviated DAGA), we arrive at a general set of ordinary differential equations modelling the dynamics of abundance, trait mean and trait variance. We discuss the relationship between DAGA and previous models of phenotypic evolution that have been employed in theoretical quantitative genetics. Noting that a natural stochastic generalization of DAGA has been previously derived, we find a path to derive SDE describing the evolutionary response to demographic stochasticity. We accomplish this in \S\ref{wnc} by discussing how a diffusion limit of a spatially structured branching process (i.e., a measure-valued branching process) leads to the natural SPDE appropriate for our study. We refer to this SPDE as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA). The diffusion limit in turn provides a rigorous method for constructing fitness functions used in models of evolutionary ecology. However, before we discuss measure-valued braching processes and their diffusion limits, we find it necessary to review some fundamental results on the calculus of white noise and introducing a set of mathematical tools based on these results. In \S\ref{wnc} and SM \S\ref{SDE_DERIV} we employ these tools to derive a system of SDE generalizing our deterministic results to account for demographic stochasticity. We find that the effect of demographic stochasticity on the evolution of mean trait and phenotypic variance characterizes the process of random genetic drift in continuous time. However, although biologically insightful, these equations remain difficult to analyze and implement numerically. In \S\ref{equations} we use an assumption of normally distributed trait values to simplify these expressions into formulae that are much easier to work with. We then account for the constraint of adaptive evolution on the amount of heritable variation in a population by extending these results via a model of imperfect inheritance. The resulting equations coincide with classical results in quantitative genetics as a special case. In \S\ref{coev} we combine the derived equations of population dynamics with classical niche theory to formulate a model of coevolution across a guild of $S$ species participating in exploitative competition along a common resource continuum. In SM \S\ref{diffuse} we apply a classical theorem on rescaled limits of BBM that allow for ecological interactions to provide a rigorous derivation. To gain biological insight, in \S\ref{dynamics} we numerically integrate our model of coevolution for $S=100$ species, tracking the dynamics of traits and abundances under scenarios of weak and strong competition. We include an account of the natural history of the simulated system and discuss the significance of demographic stochasticity for structuring ecological communities. In \S\ref{ecoevo} we provide expressions for selection gradients and competition coefficients implied by our model and use these expressions to investigate the relationship between the degree of competition and strength of coevolution. Finally, \S\ref{conclusion} concludes with a summary of accomplishments, a few remarks on the limits of this approach and future directions to incorporate more explicitly the effects of small populations, the genetic architecture of quantitative traits, feedbacks with ecosystem processes and the macroevolutionary history of interacting lineages.

# The framework

At the core of our approach is a stochastic analog of the replicator equation with mutation in continuous time and phenotypic space [@Taylor1978; @Schuster1983]. From this stochastic replicator-mutator equation we derive a system of SDE for the dynamics of abundance, mean trait and additive genetic variance of a population. Hence, our approach develops a quantitative genetic theory of evolutionary ecology. A popular alternative to quantitative genetics is the theory of adaptive dynamics. As demonstrated by Page and Nowak [-@PAGE2002] and Champagnat, Ferrière, and Méléard [-@Champagnat2006], the canonical equation of adaptive dynamics can be derived from the replicator-mutator equation. Thus, one could start from the atomic roots of our approach and pursue a stochastic adaptive dynamic theory instead. We choose the former in anticipation of an extension of our approach that explicitly models the genetic details of populations.

@@

## Deterministic dynamics \label{deterministic}

Our review begins by considering the dynamics of an asexually reproducing population in a homogeneous environment. For simplicity, we first assume individuals are haploid and carry one of $K$ alleles each with a different fitness expressed as growth rate before introducing a model involving a quantitative trait. Under these assumptions, the derivation of the evolution of allele frequencies due to natural selection can be derived from expressions of exponential growth. This, and a few related approaches, have been provided by Crow and Kimura [-@9781932846126, \S 5.3]. Specifically, denoting $\nu_i$ the abundance of individuals with allele $i$ and $m_i$ the growth rate of allele $i$ [called the Malthusian parameter in @9781932846126], we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation}
Starting from this model, we get the total abundance of the population as $N=\sum_{i=1}^K\nu_i$, the frequency of allele $i$ as $p_i=\nu_i/N$ and the mean fitness of the population as $\bar m=\sum_{i=1}^Kp_im_i$. Hence, we can employ some elementary calculus to derive the dynamics of abundance $dN/dt$ and the dynamics of allele frequencies $dp_1/dt,\dots,dp_K/dt$ as
\begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation}
\begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation}
Two important observations of these equations include 1) mean fitness $\bar m$ determines the abundance dynamics of the entire population and 2) allele $i$ will increase (decrease) in frequency if $m_i>\bar m$ ($<\bar m$). Equation (\ref{popgen}) is known in the field of evolutionary game theory as the replicator equation [@Hofbauer1998; @9780674023383; @Lion2018; @Taylor1978; @Schuster1983]. Instead of being explicitly focused on alleles, the replicator equation describes the fluctuations of relative abundances of various _types_ in a population in terms of the vital rates of each type. Using a matrix of transition rates between differing types, it is straight-forward to extend the replicator equation to include mutation, which is known as the replicator-mutator equation [@9780674023383].

Inspired by equations (\ref{first})-(\ref{popgen}), we derive an analog of the replicator-mutator equation for a continuum of types (that is, for a quantitative trait). In particular, we model a continuously reproducing population with trait values $x\in\mathbb{R}$ and an abundance density $\nu(x,t)$ that represents the amount of individuals in the population with trait value $x$ at time $t$. Hence, the abundance density satisfies $N(t)=\int_{-\infty}^{+\infty}\nu(x,t)dx$ and $p(x,t)=\nu(x,t)/N(t)$ is the relative density of trait $x$ which we also refer to as the phenotypic distribution.

>>>>

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={A White Noise Approach to Evolutionary Ecology},
            pdfauthor={Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{hyperref}
\usepackage[left]{lineno}
\usepackage{amsmath}
\linenumbers
\usepackage{float}
\usepackage{flafter}

\title{A White Noise Approach to Evolutionary Ecology}
\author{Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon}
\date{3/1/2020}

\begin{document}
\maketitle
\begin{abstract}
We derive the dynamics of the distribution of a quantitative character
and the abundance of a biological population from a stochastic partial
differential equation driven by space-time white noise. In the process
we develop a useful set of heuristics to operationalize the powerful,
but abstract theory of white noise and measure-valued stochastic
processes. This approach allows us to compute the full implications of
demographic stochasticity on phenotypic distributions and abundances of
populations. We demonstrate the utility of our approach by deriving a
quantitative genetic model of diffuse coevolution mediated by
exploitative competition for a continuum of resources. Other than trait
and abundance distributions, this model predicts interaction networks
parameterized by rates of interactions, competition coefficients, and
selection gradients. We briefly investigate the relationship between
selection gradients and competition coefficients. This illustrative
investigation suggests selection gradients can be either positively or
negatively correlated with competition coefficients depending on the
ratio of interspecific trait variation to intraspecific trait variation.
Hence, this approach contributes to the development of a synthetic
theory of evolutionary ecology by formalizing first principle
derivations of stochastic models that underlie rigorous investigations
of the relationship between feedbacks of biological processes and the
patterns of diversity they produce.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Our goal in this manuscript is to develop a rigorous, but accessible
approach to synthesize the stochastic dynamics of abundance, mean trait
and heritable variation in biological populations for the study of
theoretical evolutionary ecology. A primary aim of theoretical
evolutionary ecology is the development of mathematical approaches to
describe the evolution of populations and their interactions with both
the biotic and abiotic environments in which they are embedded. Given
this consideration, a natural scope for such an approach centers on
quantifying the abundance dynamics of populations and the evolution of
traits mediating their interactions as functions of relevant abiotic
factors. Although taking into account abundance, phenotype and
environment provides the basis for a partial understanding of the
complex nature of biological communities, a deeper understanding must
account for the effects of dispersal and the phylogeographic history of
interacting lineages (Kraft et al. 2007; Hickerson et al. 2010; Manceau,
Lambert, and Morlon 2016; McPeek 2017) along with the genetic basis of
ecologically relevant traits (Conner 2004; Fussman, Loreau, and Abrams
2007) and feedbacks between populations and the biogeochemical cycles
they ultimately depend on (Loreau 2010; Ågren and Andersson 2012). It is
therefore ideal that the development of any such mathematical approach
anticipates extensions to account for these important factors shaping
ecological communities, especially as empirical and conceptual work in
these directions continues to grow (Abdala-Roberts and Mooney 2014;
Kölzsch et al. 2015; Crutsinger 2015; Fitzpatrick et al. 2015, 2017;
Marx et al. 2017; Rudman et al. 2017; Skovmand et al. 2018; Nuland et
al. 2019; Harmon et al. 2019). Furthermore, the approach would benefit
from a stochastic component to capture the chance nature of biological
reality (Lande, Engen, and SÆther 2003; Meester et al. 2018; Mubayi et
al. 2019) and serve as a basis for the construction of statistical
methods that measure evolutionary and ecological processes occuring in
the wild. Such methods will tether theory to reality and allow for
rigorous tests of hypotheses on the structure and behavior of ecological
communities. In this paper we introduce a framework that establishes a
formal connection between the continuous-time dynamics of abundance and
quantitative traits in stochastically evolving populations. We then
demonstrate the utility of our framework through the derivation and
analysis of a model of diffuse coevolution and discuss how it can be
extended to account for the details mentioned above.

Current theoretical approaches to synthesize evolution and ecology have
capitalized on the fact that biological fitness plays a key role in
determining both sets of dynamics. While correlation of fitness and
genotype is the basis of evolution by natural selection, the mean
fitness across all individuals in a population determines the growth,
stasis or decline of abundance. In section \ref{deterministic} we review
the mathematical formalization of this connection, which has been
established in the contexts of population genetics (Crow and Kimura
1970; Roughgarden 1979), evolutionary game theory (Hofbauer and Sigmund
1998; Nowak 2006; Lion 2018), quantitative genetics (Lande 1982; Doebeli
1996; Lion 2018) and a unifying framework for these three distinct
approaches to evolutionary theory (Champagnat, Ferrière, and Méléard
2006) which includes our approach as a special case. We note here there
are two primary differences between our aims and those of Champagnat,
Ferrière, and Méléard (2006). First, instead of developing a unifying
approach to evolutionary theory, we focus on developing a stochastic
synthesis of population dynamics and quantitative genetics including
specific expressions for the dynamics of abundance, mean trait and
additive genetic variance. Second, although Champagnat, Ferrière, and
Méléard (2006) provide rigorous proofs of results relevant to our
approach, they do not communicate the nessecary tools we make use of for
deriving the dynamics of abundance, mean trait and additive genetic
variance. We aim to communicate these tools to a wide audience by
providing an intuitive, yet accurate treatment of white noise and
associated heuristics useful for deriving new stochastic models of
evolutionary ecology.

Reviewing the above accomplishments reveals a beautiful synthesis of
evolution and population ecology. However, it also reveals a gap in
theoretical approaches to incorporate the intrinsically random nature of
populations. Specifically, in theoretical quantitative genetics the
derivation of a population's response to random genetic drift is derived
in discrete time under the assumption of constant effective population
size using arguments based on properties of random samples (Lande 1976).
Though this approach conveniently mimics the formalism provided by the
Wright-Fisher model of population genetics, real population sizes
fluctuate over time. Furthermore, since these fluctuations are
themselves stochastic, it seems natural to derive expressions for the
evolutionary response to demographic stochasticity and consider how the
results relate to characterizations of random genetic drift. This can be
done in continuous time for population genetic models without too much
technical overhead, assuming a finite number of alleles (Lande, Engen,
and Sæther 2009; Parsons, Quince, and Plotkin 2010; Gomulkiewicz, Krone,
and Remien 2017). However, for populations with a continuum of types,
such as a quantitative trait, finding a formal approach to derive the
evolutionary response to demographic stochasticity has remained a vexing
mathematical challenge. Previous work integrating stochastic processes
with quantitative genetic models have profited from the simplifying
assumption that phenotypic and genetic variances remain constant over
time, yeilding analytical approximations for the joint dynamics of
abundance and the mean of a quantitative trait (Engen, Lande, and Sæther
2013). These results not only ignore the evolution genetic variance, but
they focus on environmental stochasticity and assume sufficiently large
populations so that demographic stochasticity is negligible. Here we
account for both demographic stochasticity and the evolution of additive
genetic variance by combining the calculus of white noise with results
on rescaled limits of branching Brownian motion processes (BBM) and
stochastic partial differential equations (SPDE), which are stochastic
analogs of deterministic partial differential equations (PDE). Our goal
has two components: 1) Establish a novel synthetic approach to
theoretical evolutionary ecology that provides a formal connection
between demographic stochasticity and random genetic drift in the
context of quantitative traits. 2) Communicate some useful properties of
space-time white noise, BBM and SPDE to as wide of audience as possible.
With this goal in mind we will not provide a rigorous treatment of any
of these deep subjects. Instead, we introduce a set of heuristics that
only require the basic concepts of Reimann integration, partial
differentiation and some exposure to Brownian motion and stochastic
ordinary differential equations (SDE). For a concise introduction to SDE
and Brownian motion, we recommend the primer by Evans (2014). Rigorous
treatments of SPDE and rescaled limits of BBM can be found in Walsh
(1986) and Dawson (1993) respectively.

Through accomplishing these goals this work yields five main results.
Firstly, we develop a concise, but accurate approach for understanding
SPDE and provide heuristics to empower mathematical evolutionary
ecologists to derive novel models that relax key assumptions commonly
made in combining models of evolutionary and abundance dynamics. Second,
we formally derive general SDE that express the joint dynamics of
abundance, mean trait and additive genetic variance for populations
exhibiting a wide range of potential phenotypic distributions. Third, we
simplify these SDE by approximating the abundance density of phenotypes
with a Gaussian curve and integrate the results with a model of
imperfect inheritance to find formula that generalize classical
quantitative genetic expressions. This includes an expression for
additive genetic variance analogous to the expression for the response
in mean phenotype found by Lande (1976). Fourth, following classical
niche theory we provide a rigorous mechanistic derivation of fitness for
populations experiencing resource competition and abiotic stabilizing
selection using diffusion limits of measure-valued branching processes.
This calculation provides a demonstration for performing such diffusion
limits that can be applied in other contexts. Fifth, we explore
implications of the resulting model of diffuse coevolution by
investigating the relationship between selection gradients and
competition coefficients.

We now summarize the basic structure of the paper. To provide motivation
for the stochastic equations developed later and background for our
model of coevolution, we begin with \S\ref{deterministic} by briefly
summarizing derivations of deterministic dynamics of biological
populations. Starting with a PDE (which we refer to as the Deterministic
Asexual Gaussian allelic model with Abundance dynamics, abbreviated
DAGA), we arrive at a general set of ordinary differential equations
modelling the dynamics of abundance, trait mean and trait variance. We
discuss the relationship between DAGA and previous models of phenotypic
evolution that have been employed in theoretical quantitative genetics.
Noting that a natural stochastic generalization of DAGA has been
previously derived, we find a path to derive SDE describing the
evolutionary response to demographic stochasticity. We accomplish this
in \S\ref{wnc} by discussing how a diffusion limit of a spatially
structured branching process (i.e., a measure-valued branching process)
leads to the natural SPDE appropriate for our study. We refer to this
SPDE as the Stochastic Asexual Gaussian allelic model with Abundance
dynamics (abbreviated SAGA). The diffusion limit in turn provides a
rigorous method for constructing fitness functions used in models of
evolutionary ecology. However, before we discuss measure-valued braching
processes and their diffusion limits, we find it necessary to review
some fundamental results on the calculus of white noise and introducing
a set of mathematical tools based on these results. In \S\ref{wnc} and
SM \S\ref{SDE_DERIV} we employ these tools to derive a system of SDE
generalizing our deterministic results to account for demographic
stochasticity. We find that the effect of demographic stochasticity on
the evolution of mean trait and phenotypic variance characterizes the
process of random genetic drift in continuous time. However, although
biologically insightful, these equations remain difficult to analyze and
implement numerically. In \S\ref{equations} we use an assumption of
normally distributed trait values to simplify these expressions into
formulae that are much easier to work with. We then account for the
constraint of adaptive evolution on the amount of heritable variation in
a population by extending these results via a model of imperfect
inheritance. The resulting equations coincide with classical results in
quantitative genetics as a special case. In \S\ref{coev} we combine the
derived equations of population dynamics with classical niche theory to
formulate a model of coevolution across a guild of \(S\) species
participating in exploitative competition along a common resource
continuum. In SM \S\ref{diffuse} we apply a classical theorem on
rescaled limits of BBM that allow for ecological interactions to provide
a rigorous derivation. To gain biological insight, in \S\ref{dynamics}
we numerically integrate our model of coevolution for \(S=100\) species,
tracking the dynamics of traits and abundances under scenarios of weak
and strong competition. We include an account of the natural history of
the simulated system and discuss the significance of demographic
stochasticity for structuring ecological communities. In \S\ref{ecoevo}
we provide expressions for selection gradients and competition
coefficients implied by our model and use these expressions to
investigate the relationship between the degree of competition and
strength of coevolution. Finally, \S\ref{conclusion} concludes with a
summary of accomplishments, a few remarks on the limits of this approach
and future directions to incorporate more explicitly the effects of
small populations, the genetic architecture of quantitative traits,
feedbacks with ecosystem processes and the macroevolutionary history of
interacting lineages.

\hypertarget{the-framework}{%
\section{The framework}\label{the-framework}}

At the core of our approach is a stochastic analog of the replicator
equation with mutation in continuous time and phenotypic space (Taylor
and Jonker 1978; Schuster and Sigmund 1983). From this stochastic
replicator-mutator equation we derive a system of SDE for the dynamics
of abundance, mean trait and additive genetic variance of a population.
Hence, our approach develops a quantitative genetic theory of
evolutionary ecology. A popular alternative to quantitative genetics is
the theory of adaptive dynamics. As demonstrated by Page and Nowak
(2002) and Champagnat, Ferrière, and Méléard (2006), the canonical
equation of adaptive dynamics can be derived from the replicator-mutator
equation. Thus, one could start from the atomic roots of our approach
and pursue a stochastic adaptive dynamic theory instead. We choose the
former in anticipation of an extension of our approach that explicitly
models the genetic details of populations.

@@

\hypertarget{deterministic-dynamics}{%
\subsection{\texorpdfstring{Deterministic dynamics
\label{deterministic}}{Deterministic dynamics }}\label{deterministic-dynamics}}

Our review begins by considering the dynamics of an asexually
reproducing population in a homogeneous environment. For simplicity, we
first assume individuals are haploid and carry one of \(K\) alleles each
with a different fitness expressed as growth rate before introducing a
model involving a quantitative trait. Under these assumptions, the
derivation of the evolution of allele frequencies due to natural
selection can be derived from expressions of exponential growth. This,
and a few related approaches, have been provided by Crow and Kimura
(1970, \S 5.3). Specifically, denoting \(\nu_i\) the abundance of
individuals with allele \(i\) and \(m_i\) the growth rate of allele
\(i\) (called the Malthusian parameter in Crow and Kimura 1970), we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation} Starting from this model, we get the total abundance of
the population as \(N=\sum_{i=1}^K\nu_i\), the frequency of allele \(i\)
as \(p_i=\nu_i/N\) and the mean fitness of the population as
\(\bar m=\sum_{i=1}^Kp_im_i\). Hence, we can employ some elementary
calculus to derive the dynamics of abundance \(dN/dt\) and the dynamics
of allele frequencies \(dp_1/dt,\dots,dp_K/dt\) as \begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation} \begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation} Two important observations of these equations include 1)
mean fitness \(\bar m\) determines the abundance dynamics of the entire
population and 2) allele \(i\) will increase (decrease) in frequency if
\(m_i>\bar m\) (\(<\bar m\)). Equation (\ref{popgen}) is known in the
field of evolutionary game theory as the replicator equation (Hofbauer
and Sigmund 1998; Nowak 2006; Lion 2018; Taylor and Jonker 1978;
Schuster and Sigmund 1983). Instead of being explicitly focused on
alleles, the replicator equation describes the fluctuations of relative
abundances of various \emph{types} in a population in terms of the vital
rates of each type. Using a matrix of transition rates between differing
types, it is straight-forward to extend the replicator equation to
include mutation, which is known as the replicator-mutator equation
(Nowak 2006).

Inspired by equations (\ref{first})-(\ref{popgen}), we derive an analog
of the replicator-mutator equation for a continuum of types (that is,
for a quantitative trait). In particular, we model a continuously
reproducing population with trait values \(x\in\mathbb{R}\) and an
abundance density \(\nu(x,t)\) that represents the amount of individuals
in the population with trait value \(x\) at time \(t\). Hence, the
abundance density satisfies \(N(t)=\int_{-\infty}^{+\infty}\nu(x,t)dx\)
and \(p(x,t)=\nu(x,t)/N(t)\) is the relative density of trait \(x\)
which we also refer to as the phenotypic distribution.
